 Exploratory Analysis of Red Wine Quality by Venkat Ramesh
========================================================

The Red Wine Quality dataset contains quality ratings on a scale of 0 to 10, 
accompanied by differetnt attributes of red wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(gridExtra)
library(GGally)
library(scales)
library(memisc)
library(dplyr)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
setwd('~/R/Proj6')
wq <- read.csv('wineQualityReds.csv', sep = ',')
```

Below is a snapshot of the various columns in the dataset and their datatypes:
```{r echo=FALSE, message=FALSE, warning=FALSE, Intro_to_Dataset}
dim(wq)
str(wq)
```


# Univariate Plots Section
Let's explore each of the columns individually, to begin with.

### Quality
Quality is the label column of interest that contains the integer score between
0 and 10. Here's the histogram of quality ratings in the dataset.


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Quality}
ggplot(data=wq, aes(x = quality)) + 
  geom_bar() + 
  scale_x_continuous(breaks=seq(1, 10, 1))
```

We see a nice normal distribution with most of the quality scores centered
around 5. Subsetting the data into those with quality less than 5 and greater
than 5 will help us see trends, since it also makes sense intuitively to place
the pass score for quality at, greater than 5. We also notice that this splits
our data into roughly two equal halves.


```{r echo=FALSE, message=FALSE, warning=FALSE, Generate_Quality_Class}
wq$quality.ordered = ordered(wq$quality)
wq$quality.class <- ordered(ifelse(wq$quality <= 5, 
                            "low_quality", "high_quality"),
                    levels = c("low_quality", "high_quality"))
table(wq$quality.class)
plt.facet.q.c <- facet_wrap(~quality.class, ncol = 2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Histogram}
# Helper function to generate histogram and facet it by quality class
plot_hist <- function(plt.data, plt.facet, 
                      plt.x.lab, plt.binwidth, plt.x.scale.breaks) {
  plt.x <- plt.data[[plt.x.lab]]
  plt.base <- ggplot(data=plt.data, aes(x = plt.x))
  plt.hist <- plt.base + 
              xlab(plt.x.lab) +
              geom_histogram(binwidth = plt.binwidth) 
  if (!missing(plt.x.scale.breaks)) {
    plt.hist <- plt.hist +
                scale_x_continuous(breaks = plt.x.scale.breaks)
  }
  plt.hist.faceted <- plt.hist + plt.facet
  grid.arrange(plt.hist, plt.hist.faceted,
               ncol = 1)
}
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Histogram_scale}
# Helper function to generate above histograms on a log10 x scale
plot_xlog_hist <- function(plt.data, plt.facet, plt.x.lab, plt.bins) {
  plt.x <- plt.data[[plt.x.lab]]
  plt.base <- ggplot(data=plt.data, aes(x = plt.x))
  plt.hist.xlog <- plt.base + 
              xlab(plt.x.lab) +
              geom_histogram(bins = plt.bins)  +
              scale_x_log10()
  plt.hist.xlog.faceted <- plt.hist.xlog + plt.facet
  grid.arrange(plt.hist.xlog, plt.hist.xlog.faceted,
               ncol = 1)
}
```


When we look at our attributes, we will also generate a faceted version of the
attribute histogram to get a peek ahead on the interaction of this attribute
to th quality. Since all of the columns are of decimal datatype, we will pick 
0.1 as our binwidth to be able to capture with enough granularity, unless we 
encounter a case where the range is too big or two small for this value.

If the distribution deviates significantly from normal distribution and is long
tailed, we will apply log10 transformation to the x axis to get a better 
picture.

### Alcohol
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Alcohol}
plot_hist(wq, plt.facet.q.c, "alcohol", 0.1, seq(1, 16, 1))
```

We see that low-quality group has long-tailed distribution for alcohol levels,
whereas the high-quality group appears to be more uniform. We will revisit this 
to investigate the correlation between quality and alcohol in our
bivariate analysis.

### Sulphates
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Sulphates}
plot_hist(wq, plt.facet.q.c, "sulphates", 0.1)
```
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Sulphates_Transformed}
plot_xlog_hist(wq, plt.facet.q.c, "sulphates", 50)
```

The distribution seems right skewed, but log10 transformation shows nothing
different in the high quality vs low quality plots.

### pH
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_pH}
plot_hist(wq, plt.facet.q.c, "pH", 0.1)

```

The distribution looks normal so no transformation is needed.

### Density

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_density}
plot_hist(wq, plt.facet.q.c, "density", 0.001)
```

Apart from the fact that the long tail outliers on the right of the distribution
belong to the high quality dataset, this is a again pretty normal.

### Total Sulfur Dioxide

We see from the data description that total sulfur dioxide of greather than 
50ppm, becomes evident in taste. We need to pay attention to see if this has an
effect on perceived quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_total.sulfur.dioxide}
plot_hist(wq, plt.facet.q.c, "total.sulfur.dioxide", 1, seq(0, 300, 50))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_total.sulfur.dioxide_Transformed}
plot_xlog_hist(wq, plt.facet.q.c, "total.sulfur.dioxide", 50)
```

### Free Sulfur Dioxide
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_free.sulfur.dioxide}
plot_hist(wq, plt.facet.q.c, "free.sulfur.dioxide", 1)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_free.sulfur.dioxide_Transformed}
plot_xlog_hist(wq, plt.facet.q.c, "free.sulfur.dioxide", 25)
```

We would expect free sulfur dioxide to be correlated with total sulfur dioxide.
We will examine this in the bi-variate section.

### Chlorides
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_chlorides}
plot_hist(wq, plt.facet.q.c, "chlorides", 0.01, seq(0.0, 0.7, 0.1))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_chlorides_Transformed}
plot_xlog_hist(wq, plt.facet.q.c, "chlorides", 60)
```

The distributions for cholrides are identical in both quality classes except for
the fact that the peak around mean is higher for high quality subset.

### Residual Sugar
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_residual.sugar}
plot_hist(wq, plt.facet.q.c, "residual.sugar", 0.1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_residual.sugar_Transformed}
plot_xlog_hist(wq, plt.facet.q.c, "residual.sugar", 50)
```

We notice that the long tail on the right of high quality subset, and would 
like to investigate if residual sugar has an effect on perceived quality. Also,
in general, the distribution of residual sugar has considerably far-off outliers
as evidenced in the box plot below.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_residual.sugar_boxplot}
ggplot(wq, aes( x = 1, y = residual.sugar ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' )
```                   

### Citric Acid
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_citric.acid}
plot_hist(wq, plt.facet.q.c, "citric.acid", 0.01)
```

We notice an almost trimodal distribution for citric acid in the low quality 
case, and they occur around means of 0.0, 0.25, 0.5 etc.

### Volatile acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_volatile.acidity}
plot_hist(wq, plt.facet.q.c, "volatile.acidity", 0.01)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_volatile.acidity_Transformed}
plot_xlog_hist(wq, plt.facet.q.c, "volatile.acidity", 80)
```


### Fixed Acidity


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_fixed.acidity}
plot_hist(wq, plt.facet.q.c, "fixed.acidity", 0.1)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_fixed.acidity_Transformed}
plot_xlog_hist(wq, plt.facet.q.c, "fixed.acidity", 60)
```

# Univariate Analysis

#### What is the structure of your dataset?
There are 1599 rows of wine quality data. Each row has 12 columns of numerical
attributes describing aspects of the red wine and a column of integers that
corresponds to a quality score between 0 and 10.


#### What is/are the main feature(s) of interest in your dataset?
Upon looking at Univariate plots, the following column appear to be interesting
in understanding the main feature of interest which is the quality of wine.
- Higher alcohol seems to be associated with high quality wine subset.
- Suphur dioxide dependence and correlation between total and free levels needs 
to be looked at.
- Acidity related columns such as pH, Citric Acid, fixed and volatile acidity 
need to be looked at overall acidic wines with lower pH seem to be higher 
quality .

#### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
N/A

#### Did you create any new variables from existing variables in the dataset?
We generated a quality class based on whether the quality score is greater than
5 or not. We may have to create additional factor variables from numericals for
trivariate analysis.

#### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
The distributions that were right skewed were transformed using log10 scaling.


# Bivariate Plots Section

Let's start our bivariate analysis by analyzing key columns identified earlier 
for their effects on quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Box_plot_quality}
# Helper method to plot quality across a given x column as boxplot.
plot_box_across_quality <- function(plt.data, plt.y.lab) {
  ggplot(data=plt.data, aes(y = plt.data[[plt.y.lab]], x = quality.ordered)) + 
  geom_boxplot(alpha = 0.5) + 
  geom_point(stat = 'summary', fun.y = mean, 
             shape = 8, color = 'red', size = 4) +
  xlab('quality') +
  ylab(plt.y.lab) + 
  geom_jitter(alpha = 0.3, color = 'steelblue2')
}
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Mean_Quality}
plot_mean_quality_trend <- function(plt.data, plt.x.lab, smooth.binwidth = 1) {
# Plot trend of mean quality across increases in a variable as line plot 
  ggplot(data=plt.data, aes(x = round(plt.data[[plt.x.lab]]/smooth.binwidth) * smooth.binwidth, y = quality)) + 
        geom_line(stat = 'summary', fun.y = mean) + 
        geom_smooth() +
        ylab('mean.quality') + 
        xlab(plt.x.lab)
}
```

### Quality vs Alcohol
```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Alcohol}
plot_box_across_quality(wq, 'alcohol')
```

The mean alcohol level at each quality level is denoted by 'x'. We see that 
overall higher alcohol levels are asssociated with higher quality, and 
at the highest quality all the datapoints are above 7 % alcohol.


### Quality vs Total Sulfur Dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Sulfur_Dioxide}
plot_box_across_quality(wq, 'total.sulfur.dioxide')
```


The total sulfur dioxide relation is more complicated. At low levels of 
less than 50ppm total sulfur dioxide, the datapoints are distributed across 
all quality levels whereas the datapoints with greater than 120ppm or so, are 
invariably in the 5-6 range always, indicating that at higher levels of sulfur
dioxide, its negative impact on perceived quality is more pronounced.

### Quality vs pH

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_pH}
plot_box_across_quality(wq, 'pH') 
```


Lower pH seems to be associated with higher quality indicating acidic wines are 
better tasting.

### Quality vs Volatile Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_volatile.acidity}
plot_box_across_quality(wq, 'volatile.acidity') 
```

However, volatile acidity is not favored since higher volatile acidity leads to
lower wine quality. Beyond volatile acidity value of 1, there are no more high
quality wine datapoints.

### Quality vs Residual Sugar

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_residual.sugar}
plot_box_across_quality(wq, 'residual.sugar') 
```


Residual sugar doesn't seem to be indicative of quality, which is contrary to 
intuition. Also, the IQR of residual sugar in the dataset is small, with some 
far off outliers which seem associated with low and mid-quality wines mostly.

### Scatter matrix

Now, let's look at the scatter matrix to identify correlation among attributes.

```{r echo=FALSE, message=FALSE, progress=FALSE, warning=FALSE, Pair_Plotting, fig.width= 10, fig.height = 10}
ggpairs(wq,
        lower = list(continuous = wrap("points", shape = I('.'))),
        upper = list(combo = wrap("box", outlier.shape = I('.'), size = 2)))
```

As expected, the acidity metrics are highly correlated with one another.
We also see correlations between acidity metrics and density, and between 
density and alcohol content. As expected, the sulfur dioxide levels also 
correlate well for free vs total levels.

### Free vs Total Sulfur Dioxide
```{r echo=FALSE, message=FALSE, warning=FALSE, Free_Total_SO}
ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide, color = quality.class), data = wq) + 
  geom_point() + geom_smooth(method = 'lm') 

```

The points spread out more at higher total levels and overall it appears
as though more free to total ratio might result in higher quality.

### Fixed acidity vs pH

```{r echo=FALSE, message=FALSE, warning=FALSE, Fixed_pH}
ggplot(aes(x = fixed.acidity, y = pH, color = quality.class), data = wq) + 
  geom_point() + geom_smooth(method = 'lm') 
```

There is a strong negative correlation as lower pH levels mean more acidity.

### Fixed vs Volatile Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Fixed_Volatile_Acidity}
ggplot(aes(x = fixed.acidity, y = volatile.acidity, color = quality.class), 
       data = wq) + 
  geom_point() + geom_smooth(method = 'lm') 
```

Again, we can see from this plot that high quality wines seem to have lower
volatile to fixed acidity ratios.


### Fixed acidity vs Citric Acid

```{r echo=FALSE, message=FALSE, warning=FALSE, Fixed_Citric_Acid}
ggplot(aes(x = fixed.acidity, y = citric.acid, color = quality.class), data = wq) + 
  geom_point() + geom_smooth(method = 'lm') 
```

Addition of higher levels of citric acid seem to be associated both with higher
quality and also higher fixed acidity.

### Density vs Fixed Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Acidity}
ggplot(aes(x = fixed.acidity, y = density, color = quality.class), data = wq) + 
  geom_point() + geom_smooth(method = 'lm') 
```

High quality wine seems to be associated with lower density for given 
fixed acidity.

### Density vs Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Alcohol}
ggplot(aes(x = alcohol, y = density, color = quality.class), data = wq) + 
  geom_point() + geom_smooth(method = 'lm') 
```

As expected most of the high quality wines are associated with higher alcohol 
levels and higher alcohol levels also seem to correlate with lower density.

# Bivariate Analysis

#### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
High quality wine seems to be mosly acidic (lower pH) but with low levels of 
volatile acidity. Alcohol levels are directly proportional to perceived quality.
High quality wines also seem to be less dense.

#### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
Density and fixed acidity seem to be strongly related in a directly proportional
way.Density and alcohol content are inversely proportional.

#### What was the strongest relationship you found?
Higher alcohol levels seem to be associated most strongly with higher quality.
Similarly, lower pH is also associated with higher quality.

# Multivariate Plots Section

Let's extend the visualization of relation between density and alcohol from 
previous section also adding a residual sugar class (less than median is low
sugar and greater than median is high sugar.)

```{r echo=FALSE, message=FALSE, warning=FALSE, Residual_class}
wq$residual.sugar.class <- ordered(ifelse(wq$residual.sugar <= 
                                         median(wq$residual.sugar), 
                                   "low_sugar", "high_sugar"),
                           levels = c("low_sugar", "high_sugar"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_Density_Alcohol}
ggplot(aes(x = alcohol, y = density, color = residual.sugar.class), data = wq) + 
  geom_point(alpha=0.8, size=1) +      
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='qual',
                   guide=guide_legend(title='Residual Sugar Level'))
```

Sweeter wines with higher sugar are also associated with greater density.
However, higher alcohol levels are associated with lower density.

```{r echo=FALSE, message=FALSE, warning=FALSE, Fixed_Density_Quality}
ggplot(aes(x = fixed.acidity, y = density, color = quality.ordered), data = wq) + 
  geom_point(alpha=0.8, size=1) +      
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))
```

This plot show further shows the effect of density and fixed acidity on quality.
The darker colors are invariable on the lower envelope of points showing that
for a given fixed acidity, lighter density wines are higher quality.

```{r echo=FALSE,message=FALSE, warning=FALSE,  Alcohol_pH_Quality}
ggplot(aes(x = pH, y = alcohol, color = quality.ordered), data = wq) + 
  geom_point(alpha = 0.8, size=1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))
```

In this plot, the darker points are mostly at the top, showing the effects of
two of the strongest indicators of quality that we identified, namely alcohol 
and acidity levels. This shows that hight quality wines are mostly high on
alcohol and occur on lower pH levels.


```{r echo=FALSE, message=FALSE, warning=FALSE, Volatile_alcohol_quality}
ggplot(aes(x = volatile.acidity, y = alcohol, color = quality.class), data = wq) + 
  geom_point(alpha=0.8, size=1)  +
  geom_smooth(method = "lm", se = FALSE,size=1) 
```

This shows a positive indicator of quality i.e. alcohol vs negative indicator of
quality namely volatile acidity and clearly the high quality wines are high on 
alcohol and low on volatile acidity.


# Multivariate Analysis

#### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
Lower pH levels, lower volatile acidity and higher alcohol levels strengthened
each other in improving wine quality.

#### Were there any interesting or surprising interactions between features?
Surprisingly residual sugar did not play a role in quality determination even 
though it increased density thereby potentially lowering quality rating.

#### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
N/A
------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

ggplot(aes(x = alcohol, y = density, color = residual.sugar.class), data = wq) + 
  geom_point(alpha=0.8, size=1) +      
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='qual',
                   guide=guide_legend(title='Residual Sugar Level')) +
  ggtitle('Effect of sugar and alcohol on density of wine') +
  xlab('Alcohol (%)') + 
  ylab('Density (g / cm^3)')
```

This plot is interesting because it explains the relationship of sugar and 
alcohol to density which was alluded to in wineQualityInfo.txt. Higher sugar
wines seem to have a consistent lead in density compared to wines in lower half
in residual sugar range. Alcohol on the other hand, brings down the density at 
higher levels. This also runs counter to intuition that sweeter wines would be 
better received, since we see that less dense and higher alcohol level wines are
normally preferred in this dataset.


### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(data=wq, aes(y = volatile.acidity, x = quality.ordered)) + 
  geom_boxplot(alpha = 0.5) + 
  geom_point(stat = 'summary', fun.y = mean, 
             shape = 8, color = 'red', size = 4) +
  xlab('Quality (out of 10)') +
  ylab('Volatile Acidity (acetic acid - g / dm^3)') + 
  geom_jitter(alpha = 0.3, color = 'steelblue2') + 
  ggtitle('Quality degradation with increasing volatile acidity')
```

This graph succintly captures the degradation in wine quality with increasing
volatile acidity.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = pH, y = alcohol, color = quality.ordered), data = wq) + 
  geom_point(alpha = 0.8, size=1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) + 
  ggtitle('Effect of alcohol and pH on Quality') + 
  xlab('pH') + 
  ylab('Alcohol (%)')
```

This final plot shows the remaining pieces of the trend in high quality wines
of having higher alcohol levels and lower pH levels indicating that acidic and 
more concentrated wines are preferred.

------

# Reflection

The biggest challenge in dealing with this dataset was the lack of domain 
knowledge on the account of being a teetotaler. However, exploring one variable
at a time and faceting by quality class to identify candidates for two variable 
and then faceting again during two variable analysis to identify three variable 
interactions, seem to uncover the relationships organically. Scatter matrix 
was particulary helpful as were summary stats in discovering trend.